---
title: "Phyloseq Preprocessing"
author: "Kristen James"
date: "11/2/2021"
output: html_document
---

# About

**Project:** FL100

**About:** Make Phyloseq objects with the phenotype data and the inputs from Dr. Mary Kable. 

**Inputs:** QIIME2 outputs from Dr. Mary Kable

**Outputs:** Outputs include: 

1) Processed QIIME2 inputs in R .rds and .csv formats
2) Phyloseq object with TMAO related phenotypes [PSOtmao]

**Updates:**
On 01-19-2022 I updated the chunks, "Original Data from Mary" and "Make Phyloseq object- specific TMAO variables". When MK was reading my manuscript, she noticed that I have some entries in the taxonomy table that say NA and some that say "g__" or "f__". While these may have different meanings, she suggests simply changing all cases to NA. (Another solution, you could manually differentaite "g__" to "g__ASV1" so that the glomming isn't wrong). I added code (from MK) to consistently change the NAs. Then, I created a new phyloseq object, PSO_tmao_NA. These will be added to the project branch named "update-taxonomy-table" in this GitHub project. The branches will eventually be merged to one. 

# Set Up

Load libraries. 

```{r libraries, warning=FALSE}
# Load libraries
library(phyloseq)
library(qiime2R)
library(tibble)
```

Load raw data from Dr. Mary Kable.

```{r Original Data from Mary}
## Metadata
# Metadata file from Mary Kable
metadata <- read.delim("../../data/raw/from_MaryKable/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.txt", stringsAsFactors = F)

# For qza files, make a temporary directory
dir.create("../../data/raw/from_MaryKable/Q2Rtemp")

## OTU Table
# Load the OTU count table 
countTbl <- read_qza(file = "../../data/raw/from_MaryKable/merged-table-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza", tmp = "../../data/raw/from_MaryKable/Q2Rtemp", rm = F)
names(countTbl) # look at names of the list components
countTbl$uuid # unique identifier of the artifact
countTbl$data[1:4,1:4] # the raw data (OTU table), look at a subset
dim(countTbl$data)
# Save the 'data' portion 
countData <- countTbl$data
colnames(countData)
# Note, many subject IDs have odd characters after their ID number. For example, "-X", "r", "-Z" etc. 
# These represent sample duplicates, repeats, etc. The metadata mapping file links the sampleID (XXXXr or XXXX-F10) and the subjectID (XXXX). To simplify downstream analysis, lets use the subject ID. (Can always revert if needed downstream).

# Create a copy to save in next chunk
countData_originalMK <- countData

# Now modify (and save downstream)
colnames(countData) <- gsub("[r,x]$","",colnames(countData))
colnames(countData) <- gsub("\\-.*$","",colnames(countData))
colnames(countData) # check that column names are only subjectID now; good. 

## Taxonomy Table
# Load the taxonomy table
taxonTbl <- read_qza(file = "../../data/raw/from_MaryKable/merged-taxonomy-dada2-retrained.qza", tmp = "../../data/raw/from_MaryKable/Q2Rtemp", rm = F)
names(taxonTbl)
taxonTbl$uuid
head(taxonTbl$data) # this reveals that the taxonomy table has classifications to species level (in few cases) (click the right arrow to see). 
# Make taxonomy table human readable (Nazmul's way and the original way I created the phyloseq objects)
#taxTbl<-taxonTbl$data %>% as_tibble() %>% tidyr::separate(Taxon, sep=";", c("Kingdom","Phylum","Class","Order","Family","Genus","Species"))
# View
#taxTbl

# Update on 01-19-2022
# From Mary Kable Jan.14.2022 via email - she suggests we want all unnamed taxa to be NA. For example, an entry that reads "g__" should just say NA. 
tax <- as.data.frame(taxonTbl$data) %>% dplyr::select(-Confidence) %>%
  column_to_rownames("Feature.ID") %>%
  tidyr::separate(Taxon, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";") %>%
  as.matrix()

# Trim tax to include only SVs in SV table
taxt <- tax[c(rownames(tax)%in%rownames(countData)),]
 
# Replace all empty levels with NA for the sake of consistancy
# This prevents inaccurate glomming later
taxt[,grep("Species", colnames(taxt))][grep("s__$", taxt[,grep("Species", colnames(taxt))])] <- NA
taxt[,grep("Genus", colnames(taxt))][grep("g__$", taxt[,grep("Genus", colnames(taxt))])] <- NA
taxt[,grep("Family", colnames(taxt))][grep("f__$", taxt[,grep("Family", colnames(taxt))])] <- NA
taxt[,grep("Order", colnames(taxt))][grep("o__$", taxt[,grep("Order", colnames(taxt))])] <- NA
taxt[,grep("Class", colnames(taxt))][grep("c__$", taxt[,grep("Class", colnames(taxt))])] <- NA
taxt[,grep("Phylum", colnames(taxt))][grep("p__$", taxt[,grep("Phylum", colnames(taxt))])] <- NA
```

Save

```{r Save QIIME2 files as csv and rds}
# First, save the count table (with sampleID names) - we won't plan to use this but good to have a copy
write.csv(countData_originalMK, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm.csv")
saveRDS(countData_originalMK, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm.rds")

# Save 3 items, which we will use:

# Save the metadata .txt as a csv so that it can be easily altered. 
write.csv(metadata, "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.csv")
saveRDS(metadata, "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")

# Save the count table (with subjectID names)
write.csv(countData, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.csv")
saveRDS(countData, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")

# Save the human readable taxonomy table
write.csv(taxTbl, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.csv")
saveRDS(taxTbl, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")

# Save the human readable taxonomy table with consistent NAs per MK's suggestion - 01/19/2022
write.csv(taxt, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.csv")
saveRDS(taxt, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.rds")
```

Alpha diversity statistics 

```{r Load Diversity .qza files}
md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")
rownames(md) <- md[,"Description"]

# Faith PD
FaithPD <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/faith_pd_vector.qza", tmp = "Q2RTemp", rm = F)
FaithPD_DF <- FaithPD$data #353, not 355... 

# Shannon
shannon <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/shannon_vector.qza", tmp = "Q2RTemp", rm = F)
shannon_DF <- shannon$data #353, not 355... 

# Observed ASV
Observed <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/observed_otus_vector.qza", tmp = "Q2RTemp", rm = F)
Observed_DF <- Observed$data #353, not 355... 

# Evenness ASV
evenness <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/evenness_vector.qza", tmp = "Q2RTemp", rm = F)
evenness_DF <- evenness$data #353, not 355... 

# Merge all of the alpha diversity measures by row names (or a common column name)
l = list(FaithPD_DF, shannon_DF, Observed_DF, evenness_DF)
# Make into one DF
Q2_alpha_DF <- Reduce(merge, lapply(l, function(x) data.frame(x, rn = row.names(x))))

# Now modify (and save downstream)
Q2_alpha_DF$rn <- gsub("[r,x]$","",Q2_alpha_DF$rn)
Q2_alpha_DF$rn <- gsub("\\-.*$","",Q2_alpha_DF$rn)

# Merge with MetaData
metadata2 <- merge(md, Q2_alpha_DF,
                   by.x = 0,
                   by.y = "rn",
                   all.x = TRUE)

rownames(metadata2) <- metadata2[,"Row.names"]
metadata2 <- subset(metadata2, select =-c(Row.names))

# comment out the all.x line in the merge command and run this code; this shows you that 6072 and 7063 don't have alpha stats.
#md2[!(rownames(md2) %in% metadata2$Row.names),] # 6072 and 7063 get left out... no alpha stats for them

# Save
write.csv(metadata2, file = "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.csv")
saveRDS(metadata2, file = "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.rds")

```

# Phyloseq Objects 

We want to make a phyloseq object for each type of data we are interested in. I find this easier than making one master phyloseq object since sometimes the commands don't handle NAs well. We will make new categorical columns based on the cohort's median value of a given metabolite. Once that is complete, we can create the phyloseq object and save for future use. 

## TMAO Phyloseq Object

```{r TMAO phyloseq preprocess}
# Load Mary's original memtadata file... that we converted into a csv in the above chunk. 
md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.rds")
rownames(md) <- md[,"Description"]

# Load phenotype data HERE 
df_tmao <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
colnames(df_tmao[,1:10])

# Trim df to subjects WITH TMAO (primary variable)
df_tmao <- df_tmao[which(complete.cases(df_tmao$tmao) == TRUE),]

# Take a subset
df <- df_tmao
df <- subset(df, select = c( pa_ee_total, endo_ln_rhi, endo_al75, betaine, carnitine, tmao, choline, creatinine, phosphocholine, hei_asa24_totalveg, hei_asa24_totalproteinfoods, hei_asa24_sodium, hei_asa24_refinedgrains, hei_asa24_sfat, hei_asa24_addsug, hei_asa24_wholegrain, hei_asa24_totalscore ))

# Transform
df$tmao_log <- log(df$tmao)

# First, check the structure of the data. Make sure everything is numeric/integer. 
str(df[,1:15])

# In the event that some variables are factors, run the following code. (Can run regardless, if the original is already numeric the output will be identical)
# In order to make them numeric we first have to turn them to characters. 
df1 <- df # save a copy
# This loops says, "if you are a factor, then first become a character, then a number; if you aren't a factor, stay how you are"
df1[] <- lapply(df1, function(x) {
    if(is.factor(x)) as.numeric(as.character(x)) else x
})

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
  if(is.character(x)) as.factor(x) else x
})
# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
str(df3[,1:15])

# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
df3.5 <- df3
df3.5[] <- lapply(df3, function(x) {
  relevel(x, ref = "below")
})
str(df3.5)


# Merge categorical data with numeric data
df4 <- merge(df, df3, by = 0)
rownames(df4) <- df4[,1]
df4 <- df4[,-1]

# Merge with metadata file
md2 <- merge(md, df4, by.x = "Description", by.y = 0) # Description = subjectID in the md object
# 103 subjects. These are subjects with biocrates AND microbiome data. 

# Assign 'Description' (subject ID in XXXX terms) to row names
rownames(md2) <- md2$Description

# Make pathologic variables based on TMAO
md2$TMAO_5 <- ifelse(md2$tmao <= 5, "below5", "above5")
md2$TMAO_5 <- as.factor(md2$TMAO_5)
md2$TMAO_5 <- relevel(md2$TMAO_5, ref = "below5")
# 6
md2$TMAO_6 <- ifelse(md2$tmao <= 6, "below6", "above6")
md2$TMAO_6 <- as.factor(md2$TMAO_6)
md2$TMAO_6 <- relevel(md2$TMAO_6, ref = "below6")
# 7
md2$TMAO_7 <- ifelse(md2$tmao <= 7, "below7", "above7")
md2$TMAO_7 <- as.factor(md2$TMAO_7)
md2$TMAO_7 <- relevel(md2$TMAO_7, ref = "below7")

# Lastly, add age, sex, bmi_final to dataframe
df_bins <- subset(df_tmao, select = c(sex, age, bmi_final))
# Merge
md3 <- merge(md2, df_bins, by=0)
row.names(md3) <- md3[,"Row.names"]
md3 <- subset(md3, select = -c(Row.names))
```

### Save sample data

```{r Save ln biocrates data}
write.csv(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.csv")
saveRDS(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
```

### Make Phyloseq object - specific TMAO variables

```{r Make Phyloseq object- specific TMAO variables}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
#taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds") 
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.rds") 
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

# Create phyloseq object
PSOtmao_NA <- phyloseq(
  otu_table(ASVr, taxa_are_rows = T), 
  phy_tree(treeR$data), 
  tax_table(as.data.frame(taxtable) %>% 
              #dplyr::select(-Confidence) %>% 
              #column_to_rownames("Feature.ID") %>% 
              as.matrix()), #moving the taxonomy to the way phyloseq wants it
  sample_data(metadata_a)
)
print(PSOtmao_NA)

# Save the phyloseq object
#save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData")
save(PSOtmao_NA, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_220119.RData")
```
